Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-01-06 draft.tex
Some minor corrections. Added comments about needed citations.
Some minor edits for completeness.

\newcommand{\eps}{\epsilon} 
\newcommand{\veps}{\varepsilon}

@@

We consider constraints on SIDM specifically within the context of dark electromagnetism models. 
Dark electromagnetism models are models in which a hidden, dark, sector contains a broken U(1)$'$ 
symmetry and the U(1)$'$ gauge boson is kinetically mixed with the standard model photon. For the 
purposes of this study, this is important because it demands that the Lagrangian contains terms such as 
%
\begin{equation}
\mathcal{L}_{\mathrm{int}} = g_{\chi} \bar{\chi} \, \tilde{\slashed{A'} } \chi + q \bar{f}\, \tilde{\slashed{A} } f, 
\end{equation}
%
where $\chi$ is the dark matter, $g_{\chi}$ is the dark coupling, $\tilde{A}'$ is the dark gauge boson, 
$f$ is a standard model fermion of charge $q$, and $\tilde{A}$ is the standard model gauge boson. 
The kinetic mixing causes the $\tilde{A}$ to be an admixture of the massless photon, $A$, and 
the dark photon, $A'$ of mass $m_{\mathrm{A'}}$, so that the dark matter particles are coupled to the standard model 
fermions with a coupling constant $\epsilon q$, where $\epsilon$ is the kinetic mixing parameter. 
The first term in this interaction gives rise to the dark matter self-interactions.

Dark gauge bosons are produced in astrophysical environments such as supernova 
cores via brehmsstrahlung off of standard model particles. This brehmsstrahlung 
occurs through the $\epsilon q$ coupling. The rate of brehmsstrahlung 
also depends upon the mass of the $A'$. Consequently, 
supernova cooling can constrain the mixing $\epsilon$ as a function of 
$m_{A'}$ for such models and producing such a constraint is the 
primary aim of this paper.

@@

The integration is performed numerically to obtain $ Q_i$, which is the rate of energy emission per unit volume associated with either the $pn$ or $pp$ process. To obtain the dark gauge boson luminosity, it is assumed that production takes place in a core volume $V$ of radius 1 km, giving $ \mathcal{L}_A = V(Q_{pp} + Q_{pn}) $. The luminosity $\mathcal{L}_A $ is a complicated function of $ m_A $ and $ T $, but is simply proportional to the square of the parameter $ \epsilon $.  Pulling that out of the expression gives a constraint 
\beq
\epsilon^2 I_A(m_A, T) \le \mathcal{L}_\nu
\eeq
or a bound on the coupling of 
\beq 
\epsilon \le \sqrt{\frac{4.1 \times 10^{37} MeV^2}{I_A(m_A, T)}} 
\eeq 
and the exclusion region is generated by varying the mass.

\subsection{Decay limit}
The constraint from the streaming limit above provides an upper bound on the allowed coupling (or lower bound on the excluded coupling) by considering the production on the $A'$. Naively it might be expected that this suffices, in that all higher couplings are excluded. However, in order to function as a cooling channel, enough of the produced dark gauge bosons must escape the supernova. As the coupling increases, so do processes that prevent the escape, and so the excluded region has an upper bound, above which the luminosity again drops below the neutrino luminosity. The first such limit may be found by considering decay of the dark bosons into Standard Model particles, and assuming that these SM particles are trapped in the supernova core and contribute nothing to the cooling. The dark boson has a typical lifetime of 
\beq
l = \frac{3 E_{A}}{N_{eff} m_A^2 \epsilon^2},
\eeq
and so the fraction escaping the supernova before decaying is given by 
\beq
e^{r_{decay}/l} = e^{r_{decay} N_{eff} m_A^2 \epsilon^2/(3 E_A)}.
\eeq
To take this into account, the above exponential factor is simply appended to the phase space integrand, and the calculation then proceeds as before. The limit is derived from the same equation, with the only complication being the fact that $ I_A $ is now a function of $ \epsilon $, in addition to $ m_A, T $. This makes the numerical calculation slightly more complicated, but otherwise changes nothing of importance, with now the constraint a transcendental equation,  
\beq
\epsilon \le \sqrt{\frac{4.1 \times 10^{37} MeV^2}{I_A(m_A, T, \epsilon)}}.
\eeq

@@

Having obtained an expression for $ \kappa $, we can now find the optical depth as follows: define a new quantity $ \tau_R = \kappa_R \rho_R R $. We then have 
\beq \kappa \rho R = \tau_R (\frac{\rho}{\rho_R})^2 (\frac{T_R}{T})^{3/2}  \eeq
This is combined with the expressions for the density and temperature as a function of r and plugged in to the integral expression for the optical depth to obtain 
\bea \tau_x &=& \int_{r_x}^{\infty} \tau_R (\frac{R}{r})^{3n/2} \\
 &=& \frac{\tau_r}{\frac{3n}{2}-1} (\frac{T_A}{T_R})^{(9/2-3/n)} \eea
and the bound on the coupling is finally found by requiring  $ \tau_x(\epsilon,m_A) \le 2/3 $

@@

Once the integration is complete the calculation of the trapping and streaming limits is a straightforward application of the equations previously derived, and proceeded as outlined above. The decay limit is somewhat harder, since $ \epsilon $ appears on both sides of the equation. Again we employed an unsubtle approach to solving the problem: $ \epsilon $ was set to an arbitrary value where the constraint was satisfied, and then iteratively reduced until the constraint was no longer satisfied. This obviously introduces another source of error, but with a sufficiently small interval in $ \epsilon $ this is negligible. The one further slight complication is that at after a certain value of $ M_A $ the decay limit rapidly goes to zero, at which point the procedure was terminated.

@@

\begin{figure}[th]
\includegraphics[width=9cm]{stages.eps}
\caption{The changes in the bounds generated as each stage of the calculation departs from Ref.~\cite{dent_etal12}. Interestingly the kinematics have apparently relatively little effect.
\arz{Cameron, several to-do items for this figure. On the y-axis, the symbol should be "$\epsilon$" rather than 
"e". Can we write numbers as "$10^{-9}$" rather than "1e-09" and so on? Can we make the axis labels and 
tick mark labels larger? Can you make all of the lines thicker? Can you make each of the four lines a distinct color?}
}
\end{figure}

@@

\begin{figure}[th]
	\includegraphics[width=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{dent_etal12}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\epsilon \approx  1e-8-9 $.
\arz{Cameron, some similar comments as for Figure 1. Can we write numbers as "$10^{-9}$" rather 
than "1e-09" and so on? Can we make the axis labels and tick mark labels larger?}	
	}
\end{figure}

@@

%------------ APPENDIX
\appendix
\section{Bremsstrahlung diagrams}
	Each of the $p-p-\pi $ vertices contribute a factor of $g_{pp}$,  the pseudoscalar pion-nucleon coupling. $f_{pp}$, the pseudovector pion-nucleon coupling, is related by $g_{pp} = \frac{2M_N}{m_\pi} f_{pp} $. The $ p-p-A' $ vertex contributes $e\epsilon \gamma_\mu$, where $\epsilon$ is the mixing strength between the photon and dark photon. Labeling the contributions according to the Feynman diagrams in Fig. , the matrix elements are 
	\bea
	M_1 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed{\varepsilon} u(p_1),\\
	M_2 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p3) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed \varepsilon  u(p_1), \\
	M_3 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_4 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_5 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_6 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_7 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),\\
	M_8 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \epsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p_3) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),
	\eea
	where the exchange momenta are defined by $ k = p_2-p_4 $ and $ l = p_2-p_3 $, and the dark photon's polarization is given by $\varepsilon$. 
	For $pn$ scattering, four of the diagrams are simply the same as those given above. The only new Feynman diagram comes from bremsstrahlung off the internal pion, and results in  
	 \beq
	 M'_5 = \frac{4 M_N}{ m_\pi} \frac{f_{pn}^2 e \epsilon}{l^2-m_\pi^2
	 }  \frac{1}{(l-q_a)^2 - m_\pi^2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p_3) u(p_2) (q_a - 2l)\cdot \varepsilon
	 \eeq
	 Note that these matrix elements agree with Ref.~\cite{dent_etal12}.
\acknowledgments

>>>>

\newcommand{\eps}{\varepsilon} 
\newcommand{\veps}{\varepsilon}

@@

\arz{Need to add citations to many of the Feng et al. papers here.}
We consider constraints on SIDM specifically within the context of dark electromagnetism models. 
Dark electromagnetism models are models in which a hidden, dark, sector contains a broken U(1)$'$ 
symmetry and the U(1)$'$ gauge boson is kinetically mixed with the standard model photon. For the 
purposes of this study, this is important because it demands that the Lagrangian contains terms such as 
%
\begin{equation}
\mathcal{L}_{\mathrm{int}} = g_{\chi} \bar{\chi} \, \widetilde{\slashed{A'}} \chi + q \bar{f}\, \widetilde{\slashed{A}} f, 
\end{equation}
%
where $\chi$ is the dark matter, $g_{\chi}$ is the dark coupling, $\widetilde{A'}$ is the dark gauge boson, 
$f$ is a standard model fermion of charge $q$, and $\widetilde{A}$ is the standard model gauge boson. 
The kinetic mixing, through a term 
$\frac{1}{2}\frac{\varepsilon}{\sqrt{1+\varepsilon^2}}\, \widetilde{F}_{\mu \nu}\widetilde{F}'^{\mu \nu}$ 
in the Lagrangian causes the $\widetilde{A}$ to be an admixture of the massless photon $A$, 
and the dark photon $A'$, 
of mass $m_{\mathrm{A'}} = m_{\mathrm{\widetilde{A}'}} \sqrt{1 + \varepsilon^2} \simeq m_{\widetilde{A}'}$ because 
the viable parameter range has $\varepsilon \ll 1$. The dark matter particles are thereby coupled 
to the standard model fermions with a coupling constant $\varepsilon q$, 
where $\varepsilon$ is the kinetic mixing parameter. The first term in this 
interaction Lagrangian gives rise to the dark matter self-interactions.

Dark gauge bosons are produced in astrophysical environments such as supernova 
cores primarily via brehmsstrahlung off of standard model particles. This brehmsstrahlung 
occurs through the $\varepsilon q$ coupling to charged standard model particles, 
in this  particular case the proton and pion. The rate of brehmsstrahlung 
depends upon both $\varepsilon$ and the mass of the $A'$. Consequently, 
supernova cooling can constrain the mixing $\varepsilon$ as a function of 
$m_{A'}$ for such models. Delineating such a constraint is the 
primary aim of this paper.

@@

The integration is performed numerically to obtain $ Q_i$, which is the rate of energy emission per unit volume associated with either the $pn$ or $pp$ process. To obtain the dark gauge boson luminosity, it is assumed that production takes place in a core volume $V$ of radius 1 km, giving $ \mathcal{L}_A = V(Q_{pp} + Q_{pn}) $. The luminosity $\mathcal{L}_A $ is a complicated function of $ m_A $ and $ T $, but is simply proportional to the square of the parameter $ \varepsilon $.  Pulling that out of the expression gives a constraint 
\beq
\varepsilon^2 I_A(m_A, T) \le \mathcal{L}_\nu
\eeq
or a bound on the coupling of 
\beq 
\varepsilon \le \sqrt{\frac{4.1 \times 10^{37} MeV^2}{I_A(m_A, T)}} 
\eeq 
and the exclusion region is generated by varying the mass.

\subsection{Decay limit}
The constraint from the streaming limit above provides an upper bound on the allowed coupling (or lower bound on the excluded coupling) by considering the production on the $A'$. Naively it might be expected that this suffices, in that all higher couplings are excluded. However, in order to function as a cooling channel, enough of the produced dark gauge bosons must escape the supernova. As the coupling increases, so do processes that prevent the escape, and so the excluded region has an upper bound, above which the luminosity again drops below the neutrino luminosity. The first such limit may be found by considering decay of the dark bosons into Standard Model particles, and assuming that these SM particles are trapped in the supernova core and contribute nothing to the cooling. The dark boson has a typical lifetime of 
\beq
l = \frac{3 E_{A}}{N_{eff} m_A^2 \varepsilon^2},
\eeq
and so the fraction escaping the supernova before decaying is given by 
\beq
e^{r_{decay}/l} = e^{r_{decay} N_{eff} m_A^2 \varepsilon^2/(3 E_A)}.
\eeq
To take this into account, the above exponential factor is simply appended to the phase space integrand, and the calculation then proceeds as before. The limit is derived from the same equation, with the only complication being the fact that $ I_A $ is now a function of $ \varepsilon $, in addition to $ m_A, T $. This makes the numerical calculation slightly more complicated, but otherwise changes nothing of importance, with now the constraint a transcendental equation,  
\beq
\varepsilon \le \sqrt{\frac{4.1 \times 10^{37} MeV^2}{I_A(m_A, T, \varepsilon)}}.
\eeq

@@

Having obtained an expression for $ \kappa $, we can now find the optical depth as follows: define a new quantity $ \tau_R = \kappa_R \rho_R R $. We then have 
\beq \kappa \rho R = \tau_R (\frac{\rho}{\rho_R})^2 (\frac{T_R}{T})^{3/2}  \eeq
This is combined with the expressions for the density and temperature as a function of r and plugged in to the integral expression for the optical depth to obtain 
\bea \tau_x &=& \int_{r_x}^{\infty} \tau_R (\frac{R}{r})^{3n/2} \\
 &=& \frac{\tau_r}{\frac{3n}{2}-1} (\frac{T_A}{T_R})^{(9/2-3/n)} \eea
and the bound on the coupling is finally found by requiring  $ \tau_x(\varepsilon,m_A) \le 2/3 $

@@

Once the integration is complete the calculation of the trapping and streaming limits is a straightforward application of the equations previously derived, and proceeded as outlined above. The decay limit is somewhat harder, since $ \varepsilon $ appears on both sides of the equation. Again we employed an unsubtle approach to solving the problem: $ \varepsilon $ was set to an arbitrary value where the constraint was satisfied, and then iteratively reduced until the constraint was no longer satisfied. This obviously introduces another source of error, but with a sufficiently small interval in $ \varepsilon $ this is negligible. The one further slight complication is that at after a certain value of $ M_A $ the decay limit rapidly goes to zero, at which point the procedure was terminated.

@@

\begin{figure}[th]
\includegraphics[width=9cm]{stages.eps}
\caption{The changes in the bounds generated as each stage of the calculation departs from Ref.~\cite{dent_etal12}. Interestingly the kinematics have apparently relatively little effect.
\arz{Cameron, several to-do items for this figure. On the y-axis, the symbol should be "$\varepsilon$" rather than 
"e". Can we write numbers as "$10^{-9}$" rather than "1e-09" and so on? Can we make the axis labels and 
tick mark labels larger? Can you make all of the lines thicker? Can you make each of the four lines a distinct color?}
}
\end{figure}

@@

\begin{figure}[th]
	\includegraphics[width=9cm]{endtoend.png}
	\caption{A comparison of the final result for the excluded region with the result from Ref.~\cite{dent_etal12}. The excluded region is clearly significantly less restrictive, and opens up the region of parameter space for $\varepsilon \approx  1e-8-9 $.
\arz{Cameron, some similar comments as for Figure 1. Can we write numbers as "$10^{-9}$" rather 
than "1e-09" and so on? Can we make the axis labels and tick mark labels larger?}	
	}
\end{figure}

@@

%------------ APPENDIX
\appendix
\section{Bremsstrahlung diagrams}
	Each of the $p-p-\pi $ vertices contribute a factor of $g_{pp}$,  the pseudoscalar pion-nucleon coupling. $f_{pp}$, the pseudovector pion-nucleon coupling, is related by $g_{pp} = \frac{2M_N}{m_\pi} f_{pp} $. The $ p-p-A' $ vertex contributes $e\varepsilon \gamma_\mu$, where $\varepsilon$ is the mixing strength between the photon and dark photon. Labeling the contributions according to the Feynman diagrams in Fig. , the matrix elements are 
	\bea
	M_1 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed{\varepsilon} u(p_1),\\
	M_2 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_1} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p3) \gamma_5 (\slashed{p}_1 - q_a +M_N)\slashed \varepsilon  u(p_1), \\
	M_3 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_4 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 - 2 q_a \cdot p_2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \gamma_5 (\slashed{p}_2 - q_a +M_N)\slashed \varepsilon  u(p_2), \\
	M_5 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p3) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_6 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_3} \bar{u}(p_3) \gamma_5 u(p_1) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_3 + q_a +M_N)\gamma_5 u(p_2),\\
	M_7 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{l^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_3) \gamma_5 u(p_2) \bar{u}(p4) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),\\
	M_8 &=& \frac{4 M_N}{ m_\pi} \frac{f_{pp}^2 e \varepsilon}{k^2-m_\pi^2
	}  \frac{1}{m_{A^\prime}^2 + 2 q_a \cdot p_4} \bar{u}(p_4) \gamma_5 u(p_2) \bar{u}(p_3) \slashed{\varepsilon} (\slashed{p}_4 + q_a +M_N)\gamma_5 u(p_1),
	\eea
	where the exchange momenta are defined by $ k = p_2-p_4 $ and $ l = p_2-p_3 $, and the dark photon's polarization is given by $\varepsilon$. 
	For $pn$ scattering, four of the diagrams are simply the same as those given above. The only new Feynman diagram comes from bremsstrahlung off the internal pion, and results in  
	 \beq
	 M'_5 = \frac{4 M_N}{ m_\pi} \frac{f_{pn}^2 e \varepsilon}{l^2-m_\pi^2
	 }  \frac{1}{(l-q_a)^2 - m_\pi^2} \bar{u}(p_4) \gamma_5 u(p_1) \bar{u}(p_3) u(p_2) (q_a - 2l)\cdot \varepsilon
	 \eeq
	 Note that these matrix elements agree with Ref.~\cite{dent_etal12}.
\acknowledgments
